---
title: "R Day 2"
author: "Andrew + Alathea"
date: "29 September 2015"
output: html_document
---

```{r load_gapminder, include=FALSE}
source("../r-novice-gapminder-1/tools/chunk-options.R")
# Silently load in the data so the rest of the lesson works
gapminder <- read.csv("../r-novice-gapminder-1/data/gapminder-FiveYearData.csv", header=TRUE)
```

## What we will do today [Andrew]

* A quick review of subsetting
* A more extensive review of plots
* Learning how to manipulate data frames
* Functions
* Saving plots and data

## Organization [Andrew]

Create a folder for `data`, `cleaned_data`, `scripts`, `figures`, `manuscript`

## Review data frames and plotting [Alathea]

### Methods of subsetting

1. Single dimension square brackets `x[a]`
2. Multi-dimension square brackets `x[a, b]`
3. Names `x["a"]` or `x["a",]`
4. Double square brackets `x[[a]]`
5. Dollar sign `x$a`

#### Which data structures can each subsetting method be used on?

### More complex subsetting

1. Selecting items with certain names `%in%
2. Selecting items by comparison `>`, `<`, `==`

### Plots

Reminder of the example from yesterday:

```{r lifeExp-vs-gdpPercap-scatter, message=FALSE}
library(ggplot2)
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point()
```

* ggplot works by adding layers to plots
* geoms add new layers of different types (e.g. points, lines, bars)

By itself, the call to `ggplot` isn't enough to draw a figure:

> #### Challenge 2 {.challenge}
>
> In the previous examples and challenge we've used the `aes` function to tell
> the scatterplot **geom** about the **x** and **y** locations of each point.
> Another *aesthetic* property we can modify is the point *color*. Modify the
> code from the previous challenge to **color** the points by the "continent"
> column. What trends do you see in the data? Are they what you expected?
>

```{r lifeExp-vs-gdpPercap-scatter2, message=FALSE}
library(ggplot2)
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap, color = continent)) +
  geom_point()
```


### Layers

Using a scatterplot probably isn't the best for visualising change over time.
Instead, let's tell `ggplot` to visualise the data as a line plot:

```{r lifeExp-line}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country, color=continent)) +
  geom_line()
```

Instead of adding a `geom_point` layer, we've added a `geom_line` layer. We've
added the **by** *aesthetic*, which tells `ggplot` to draw a line for each
country.

But what if we want to visualise both lines and points on the plot? We can
simply add another layer to the plot:

```{r lifeExp-line-point}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country, color=continent)) +
  geom_line() + geom_point()
```

It's important to note that each layer is drawn on top of the previous layer. In
this example, the points have been drawn *on top of* the lines. Here's a
demonstration:

```{r lifeExp-layer-example-1}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country)) +
  geom_line(aes(color=continent)) + geom_point()
```

In this example, the *aesthetic* mapping of **color** has been moved from the
global plot options in `ggplot` to the `geom_line` layer so it no longer applies
to the points. Now we can clearly see that the points are drawn on top of the
lines.

> #### Challenge 3 {.challenge}
>
> Switch the order of the point and line layers from the previous example. What
> happened?
>

### Transformations and statistics

Ggplot also makes it easy to overlay statistical models over the data. To
demonstrate we'll go back to our first example:

```{r lifeExp-vs-gdpPercap-scatter3, message=FALSE}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap, color=continent)) +
  geom_point()
```

Currently it's hard to see the relationship between the points due to some strong
outliers in GDP per capita. We can change the scale of units on the y axis using
the *scale* functions. These control the mapping between the data values and
visual values of an aesthetic.

> #### Challenge 4 {.challenge}
>
> How could you modify the plot to have a log scale on the y-axis?
>

```{r axis-scale}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() + scale_y_log10()
```

We can fit a simple relationship to the data by adding another layer,
`geom_smooth` (or `stat_smooth`):

```{r lm-fit}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() + scale_y_log10() + geom_smooth(method="lm")
```

We can make the line thicker by *setting* the **size** aesthetic in the
`geom_smooth` layer:

```{r lm-fit2}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() + scale_y_log10() + geom_smooth(method="lm", size=1.5)
```

There are two ways an *aesthetic* can be specified. Here we *set* the **size**
aesthetic by passing it as an argument to `geom_smooth`. Previously in the
lesson we've used the `aes` function to define a *mapping* between data
variables and their visual representation.

> #### Challenge 5 {.challenge}
>
> Modify the color and size of the points on the point layer in the previous
> example.
>
> Hint: do not use the `aes` function.
>

### Multi-panel figures

Earlier we visualised the change in life expectancy over time across all
countries in one plot. Alternatively, we can split this out over multiple panels
by adding a layer of **facet** panels:

```{r facet}
gapminder2 <- gapminder[gapminder$year > 1990,]

ggplot(data = gapminder2, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() + scale_y_log10() + geom_smooth(method="lm", size=1.5) +
  facet_wrap(~ year)
```

The `facet_wrap` layer took a "formula" as its argument, denoted by the tilde
(~). This tells R to draw a panel for each unique value in the country column
of the gapminder dataset.

### Modifying text

To clean this figure up for a publication we need to change some of the text
elements.

We can do this by adding a couple of different layers.

```{r theme}
ggplot(data = gapminder2, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() + scale_y_log10() + geom_smooth(method="lm", size=1.5) +
  facet_wrap(~ year) +
  xlab("Life expectancy") + ylab("GDP per capita") + ggtitle("Figure 1")
```


This is just a taste of what you can do with `ggplot2`. RStudio provides a
really useful [cheat sheet][cheat] of the different layers available, and more
extensive documentation is available on the [ggplot2 website][ggplot-doc].
Finally, if you have no idea how to change something, a quick google search will
usually send you to a relevant question and answer on Stack Overflow with reusable
code to modify!

[cheat]: http://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf
[ggplot-doc]: http://docs.ggplot2.org/current/


> #### Challenge 6 {.challenge}
>
> Create a density plot of GDP per capita, filled by continent.
>
> Advanced:
>  - Transform the x axis to better visualise the data spread.
>  - Add a facet layer to panel the density plots by year.
>

## Working with data frames [Alathea]

Talk about:

* Adding columns to data frames using calculations

### Create a new column

You can add a column to a data frame using the `$` that we used for subsetting.

```{r}
iris$Family <- "Iridaceae"
```

What if we want to add a column based on values in the other column?

For example, we may want to know the area of iris petals using their length and width.

You can repeat a process on every row in a data frame.  This time you must tell R that you are using values from within the data.frame.

```{r petal_area}
iris$Petal.Area <- iris$Petal.Length * iris$Petal.Width
head(iris)
```

This code can be simplified a little with the `with` command:

```{r with}
iris$Petal.Area <- with(iris, Petal.Length * Petal.Width)
```

### Summary Stats

Another thing you may want to do with data from a data frame, is summary statistics.  For example, you could take the mean and standard deviation of petal widths:

```{r mean_petal_width}
mean_petal_width <- mean(iris$Petal.Width)
mean_petal_width

sd_petal_width <- sd(iris$Petal.Width)
sd_petal_width
```

Using both mean and standard deviation, we can also get the coefficient of variation of petal widths (sd / mean)

```{r cov_petal_width}
cov_petal_width <- sd_petal_width/mean_petal_width
cov_petal_width
```

Now, what if you want the coefficient of variation for other measures, such as petal length, sepal length, etc.?

> #### Challenge 1 {.challenge}
>
> Find the coefficient of variation (standard deviation / mean) for the variable Sepal.Length
> in the iris data set.
>

To solve the last challenge, you had to repeat the steps of taking the mean and standard deviation.  It is annoying to have to write the whole process out every time you want to know the coefficient of variation.  Wouldn't it be nice if R had a function to get this value automatically?  Luckily, you can write your own functions, which is what Andrew will talk about now!

## Functions [Andrew]

If we only had one data set to analyze, it would probably be faster to load the
file into a spreadsheet and use that to plot simple statistics. However, the
gapminder data is updated periodically, and we may want to pull in that new
information later and re-run our analysis again. We may also obtain similar data
from a different source in the future.

In this lesson, we'll learn how to write a function so that we can repeat
several operations with a single command.

> ## What is a function? {.callout}
>
> Functions gather a sequence of operations into a whole, preserving it for ongoing use. Functions provide:
>
> * a name we can remember and invoke it by
> * relief from the need to remember the individual operations
> * a defined set of inputs and expected outputs
> * rich connections to the larger programming environment
>
> As the basic building block of most programming languages, user-defined functions constitute "programming" as much as any single abstraction can. If you have written a function, you are a computer programmer.
>


### Defining a function

Let's open a new R script file in the `functions/` directory and call it functions-lesson.R.

```{r}
my_sum <- function(a, b) {
  the_sum <- a + b
  return(the_sum)
}
```

Let’s define a function fahr_to_kelvin that converts temperatures from Fahrenheit to Kelvin:

```{r}
fahr_to_kelvin <- function(temp) {
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}
```

We define `fahr_to_kelvin` by assigning it to the output of `function`.
The list of argument names are contained within parentheses.
Next, the [body](reference.html#function-body) of the function--the statements that are executed when it runs--is contained within curly braces (`{}`).
The statements in the body are indented by two spaces.
This makes the code easier to read but does not affect how the code operates.

When we call the function, the values we pass to it are assigned to those variables so that we can use them inside the function.
Inside the function, we use a [return statement](reference.html#return-statement) to send a result back to whoever asked for it.

> ## Tip {.callout}
>
> One feature unique to R is that the return statement is not required.
> R automatically returns whichever variable is on the last line of the body
> of the function. Since we are just learning, we will explicitly define the
> return statement.

Let's try running our function.
Calling our own function is no different from calling any other function:

```{r}
# freezing point of water
fahr_to_kelvin(32)
```

```{r}
# boiling point of water
fahr_to_kelvin(212)
```

> #### Challenge 1 {.challenge}
>
> Write a function called `kelvin_to_celsius` that takes a temperature in Kelvin
> and returns that temperature in Celsius
>
> Hint: To convert from Kelvin to Celsius you minus 273.15
>

#### Combining functions

The real power of functions comes from mixing, matching and combining them
into ever large chunks to get the effect we want.

Let's define two functions that will convert temperature from Fahrenheit to
Kelvin, and Kelvin to Celsius:

```{r}
fahr_to_kelvin <- function(temp) {
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}

kelvin_to_celsius <- function(temp) {
  celsius <- temp - 273.15
  return(celsius)
}
```

> #### Challenge 2 {.challenge}
>
> Define the function to convert directly from Fahrenheit to Celsius,
> by reusing the two functions above (or using your own functions if you prefer).
>


We're going to define
a function that calculates the Gross Domestic Product of a nation from the data
available in our dataset:

```{r}
# Takes a dataset and multiplies the population column
# with the GDP per capita column.
calcGDP <- function(dat) {
  gdp <- dat$pop * dat$gdpPercap
  return(gdp)
}
```

We define `calcGDP` by assigning it to the output of `function`.
The list of argument names are contained within parentheses.
Next, the body of the function -- the statements executed when you
call the function -- is contained within curly braces (`{}`).

We've indented the statements in the body by two spaces. This makes
the code easier to read but does not affect how it operates.

When we call the function, the values we pass to it are assigned
to the arguments, which become variables inside the body of the
function.

Inside the function, we use the `return` function to send back the
result. This return function is optional: R will automatically
return the results of whatever command is executed on the last line
of the function.


```{r}
calcGDP(head(gapminder))
```

That's not very informative. Let's add some more arguments so we can extract
that per year and country.

```{r}
# Takes a dataset and multiplies the population column
# with the GDP per capita column.
calcGDP <- function(dat, year=NULL, country=NULL) {
  if(!is.null(year)) {
    dat <- dat[dat$year %in% year, ]
  }
  if (!is.null(country)) {
    dat <- dat[dat$country %in% country,]
  }
  gdp <- dat$pop * dat$gdpPercap

  new <- cbind(dat, gdp=gdp)
  return(new)
}
```

If you've been writing these functions down into a separate R script
(a good idea!), you can load in the functions into our R session by using the
`source` function:

```{r, eval=FALSE}
source("functions/functions-lesson.R")
```

Ok, so there's a lot going on in this function now. In plain English,
the function now subsets the provided data by year if the year argument isn't
empty, then subsets the result by country if the country argument isn't empty.
Then it calculates the GDP for whatever subset emerges from the previous two steps.
The function then adds the GDP as a new column to the subsetted data and returns
this as the final result.
You can see that the output is much more informative than just getting a vector of numbers.

Let's take a look at what happens when we specify the year:

```{r}
head(calcGDP(gapminder, year=2007))
```

Or for a specific country:

```{r}
calcGDP(gapminder, country="Australia")
```

Or both:

```{r}
calcGDP(gapminder, year=2007, country="Australia")
```

Let's walk through the body of the function:

``` {r, eval=FALSE}
calcGDP <- function(dat, year=NULL, country=NULL) {
```

Here we've added two arguments, `year`, and `country`. We've set
*default arguments* for both as `NULL` using the `=` operator
in the function definition. This means that those arguments will
take on those values unless the user specifies otherwise.

```{r, eval=FALSE}
  if(!is.null(year)) {
    dat <- dat[dat$year %in% year, ]
  }
  if (!is.null(country)) {
    dat <- dat[dat$country %in% country,]
  }
```

Here, we check whether each additional argument is set to `null`,
and whenever they're not `null` overwrite the dataset stored in `dat` with
a subset given by the non-`null` argument.

I did this so that our function is more flexible for later. We
can ask it to calculate the GDP for:

 * The whole dataset;
 * A single year;
 * A single country;
 * A single combination of year and country.

By using `%in%` instead, we can also give multiple years or countries
to those arguments.

> #### Tip: Pass by value {.callout}
>
> Functions in R almost always make copies of the data to operate on
> inside of a function body. When we modify `dat` inside the function
> we are modifying the copy of the gapminder dataset stored in `dat`,
> not the original variable we gave as the first argument.
>
> This is called "pass-by-value" and it makes writing code much safer:
> you can always be sure that whatever changes you make within the
> body of the function, stay inside the body of the function.
>

> #### Tip: Function scope {.callout}
>
> Another important concept is scoping: any variables (or functions!) you
> create or modify inside the body of a function only exist for the lifetime
> of the function's execution. When we call `calcGDP`, the variables `dat`,
> `gdp` and `new` only exist inside the body of the function. Even if we
> have variables of the same name in our interactive R session, they are
> not modified in any way when executing a function.
>

```{r, eval=FALSE}
  gdp <- dat$pop * dat$gdpPercap
  new <- cbind(dat, gdp=gdp)
  return(new)
}
```

Finally, we calculated the GDP on our new subset, and created a new
data frame with that column added. This means when we call the function
later we can see the context for the returned GDP values,
which is much better than in our first attempt where we just got a vector of numbers.

> #### Challenge 3 {.challenge}
>
> The `paste` function can be used to combine text together, e.g:
>
> ```{r}
> best_practice <- c("Write", "programs", "for", "people", "not", "computers")
> paste(best_practice, collapse=" ")
> ```
>
>  Write a function called `fence` that takes two vectors as arguments, called
> `text` and `wrapper`, and prints out the text wrapped with the `wrapper`:
>
> ```{r, eval=FALSE}
> fence(text=best_practice, wrapper="***")
> ```
>
> *Note:* the `paste` function has an argument called `sep`, which specifies the
> separator between text. The default is a space: " ". The default for `paste0`
> is no space "".
>

> ## Tip {.callout}
>
> R has some unique aspects that can be exploited when performing
> more complicated operations. We will not be writing anything that requires
> knowledge of these more advanced concepts. In the future when you are
> comfortable writing functions in R, you can learn more by reading the
> [R Language Manual][man] or this [chapter][] from
> [Advanced R Programming][adv-r] by Hadley Wickham. For context, R uses the
> terminology "environments" instead of frames.

[man]: http://cran.r-project.org/doc/manuals/r-release/R-lang.html#Environment-objects
[chapter]: http://adv-r.had.co.nz/Environments.html
[adv-r]: http://adv-r.had.co.nz/


> #### Tip: Testing and documenting {.callout}
>
> It's important to both test functions and document them:
> Documentation helps you, and others, understand what the
> purpose of your function is, and how to use it, and its
> important to make sure that your function actually does
> what you think.
>
> When you first start out, your workflow will probably look a lot
> like this:
>
>  1. Write a function
>  2. Comment parts of the function to document its behaviour
>  3. Load in the source file
>  4. Experiment with it in the console to make sure it behaves
>     as you expect
>  5. Make any necessary bug fixes
>  6. Rinse and repeat.
>
> Formal documentation for functions, written in separate `.Rd`
> files, gets turned into the documentation you see in help
> files. The [roxygen2][] package allows R coders to write documentation alongside
> the function code and then process it into the appropriate `.Rd` files.
> You will want to switch to this more formal method of writing documentation
> when you start writing more complicated R projects.
>
> Formal automated tests can be written using the [testthat][] package.

[roxygen2]: http://cran.r-project.org/web/packages/roxygen2/vignettes/rd.html
[testthat]: http://r-pkgs.had.co.nz/tests.html

## Challenge solutions

> #### Solution to challenge 1 {.challenge}
>
> Write a function called `kelvin_to_celsius` that takes a temperature in Kelvin
> and returns that temperature in Celsius
>
> ```{r}
> kelvin_to_celsius <- function(temp) {
>  celsius <- temp - 273.15
>  return(celsius)
> }
> ```


> #### Solution to challenge 2 {.challenge}
>
> Define the function to convert directly from Fahrenheit to Celsius,
> by reusing these two functions above
>
>
> ```{r}
> fahr_to_celsius <- function(temp) {
>   temp_k <- fahr_to_kelvin(temp)
>   result <- kelvin_to_celsius(temp_k)
>   return(result)
> }
> ```
>


> #### Solution to challenge 3 {.challenge}
>
>  Write a function called `fence` that takes two vectors as arguments, called
> `text` and `wrapper`, and prints out the text wrapped with the `wrapper`:
>
> ```{r}
> fence <- function(text, wrapper){
>   text <- c(wrapper, text, wrapper)
>   result <- paste(text, collapse = " ")
>   return(result)
> }
> best_practice <- c("Write", "programs", "for", "people", "not", "computers")
> fence(text=best_practice, wrapper="***")
> ```

## Saving plots and data [Alathea]

### Saving plots

You have already seen how to save the most recent plot you create in `ggplot2`,
using the command `ggsave`. As a refresher:

```{r, eval=FALSE}
ggsave("My_most_recent_plot.pdf")
```

You can save a plot from within RStudio using the 'Export' button
in the 'Plot' window. This will give you the option of saving as a
.pdf or as .png, .jpg or other image formats.

Sometimes you will want to save plots without creating them in the
'Plot' window first. Perhaps you want to make a pdf document with
multiple pages: each one a different plot, for example. Or perhaps
you're looping through multiple subsets of a file, plotting data from
each subset, and you want to save each plot, but obviously can't stop
the loop to click 'Export' for each one.

Each new plot will show up as a new page before you turn off the device.

In this case you can use a more flexible approach. The function
`pdf` creates a new pdf device. You can control the size and resolution
using the arguments to this function.

```{r, eval=FALSE}
pdf("Life_Exp_vs_time.pdf", width=12, height=4)
ggplot(data=gapminder, aes(x=year, y=lifeExp, colour=country)) +
  geom_line()

# You then have to make sure to turn off the pdf device!

dev.off()
```

Open up this document and have a look.

> #### Challenge 1 {.challenge}
>
> Rewrite your 'pdf' command to print a second
> page in the pdf, showing a facet plot (hint: use `facet_grid`)
> of the same data with one panel per continent.
>

The commands `jpeg`, `png` etc. are used similarly to produce
documents in different formats.

### Writing data

At some point, you'll also want to write out data from R.

We can use the `write.table` function for this, which is
very similar to `read.table` from before.

Let's create a data-cleaning script, for this analysis, we
only want to focus on the gapminder data for Australia:

```{r}
library(gapminder)

aust_subset <- gapminder[gapminder$country == "Australia",]

write.table(aust_subset,
  file="gapminder-aus.csv",
  sep=","
)
```

Now open this file in a text editor and take a look.

Hmm, that's not quite what we wanted. Where did all these
quotation marks come from? Also the row numbers are
meaningless.

Let's look at the help file to work out how to change this
behaviour.

```{r, eval=FALSE}
?write.table
```

By default R will wrap character vectors with quotation marks
when writing out to file. It will also write out the row and
column names.

Let's fix this:

```{r}
write.table(
  gapminder[gapminder$country == "Australia",],
  file="gapminder-aus.csv",
  sep=",", quote=FALSE, row.names=FALSE
)
```

Now lets look at the data again 

That looks better!

> #### Challenge 2 {.challenge}
>
> Write a data-cleaning script file that subsets the gapminder
> data to include only data points collected since 1990.
>
> Use this script to write out the new subset to a file
> in the `cleaned-data/` directory.
>

```{r, echo=FALSE}
# We remove after rendering the lesson, because we don't want this in the lesson
# repository
unlink("cleaned-data", recursive=TRUE)
```
